The natural history of scavenging in vertebrates

### 2016-03-16 natural history of scavenging.tex
section on temp

\section*{Introduction}
%TG: it might be better to reorganise the intro as:
%§1 - scavenging is understudied because it's hard to say who's a scavenger
%§2 - it's even more problematic in the fossil record but people have tried some indirect approaches
%§3 - good thing is that we can apply these indirect approaches to extant systems
%§4 - in this review we discuss some stuff.
%TG: or something along these lines, obviously flexible!
Historically, scavengers have not been viewed as the most charismatic of animals. This may go some way to explaining the gap in our knowledge of the prevalence of this behaviour. Consider Professor Sanborn Tenney writing in 1877 for The American Naturalist who had this to say about one well known group, “Prominent among the mammalian scavengers are the hyenas, the ugliest in their general appearance of all the flesh eaters.” He contrasts these with “nobler kinds” of carnivores such as lions and tigers \citep{tenney1877few}. Even aside from our own subjective biases, scavenging is a difficult behaviour to detect after the fact. Without catching a carnivore in the act of killing we are left to infer how the prey was killed. Some simple heuristics can inform us, for instance, in cases where the prey item was simply too large to have been killed by the ostensible predator \citep{pobiner2008paleoecological}. But clearly, a scavenger doesn’t only feed on animals too big for it to have hunted. The obvious lack of direct behavioural data compounds the difficulty of discerning scavenging among extinct forms. Indeed, a single species of dinosaur notwithstanding, a synthesis describing the natural history of scavengers is absent from the literature. Fortunately, research on scavenging is on the rise \citep{manga2006vulture}. As a result, we are now beginning to realise the extent of this behaviour such that, “in some ecosystems, vertebrates have been documented to assimilate as much as 90\% of the available carrion" \citep{benbow2015introduction}. Even Tenney’s noble big cats are now known to take in a significant portion of carrion in their diet where some lion populations get over 50\% of their meat from carcasses. By recognising the difficulty in directly observing scavenging, a suite of methods have been used to discern the most suitable morphologies, physiologies and environments for a scavenging lifestyle to prosper. Here we chart the natural history of vertebrate scavenging on air, land and in water by focussing on the dominant vertebrate groups. %It is our aim in this review to employ these methods to gain an understanding of scavengers past and present. 
\subsection*{The Difficulty of Scavenging}
The chief hurdle to scavenging is finding a sufficient quantity of food, the occurence of which is difficult to predict in space and time. Once found the scavenger has to overcome the agents of decay produced by the action of microorganisms on the carcass. The idea of scrounging from predator kills is undermined from studies showing that in the  majority of ecosystems more animals die from disease and starvation than predation \citep{benbow2015introduction}. Thus, any animal existing as a scavenger must maximise its detection capabilities and minimise its locomotory costs \citep{ruxton2004obligate}. The habitat must also be productive enough to sustain an animal biomass that will eventually produce carcasses.

@@

Vultures represent the best known scavengers on Earth. These birds consist of two convergent groups, from the old and the new world and represent the only example of obligate vertebrate scavengers today. Given their unique position, they have been extensively studied to determine what adaptations they possess that allows them to so flourish in this niche. As such, we can begin by exploring the adapdations and the environments of vultures to draw comparisons with other scavenging species and \textit{their} environments.

@@

Aside from sight, three species within the new world family Cathartidae, (genus \textit{Cathartes}), have well developed olfactory systems \citep{AR:AR22815}. Among them are the Turkey Vultures \textit{Cathartes aura} which were able to locate 90\% of baits set out in a tropical forest \citep{houston1986olfaction}. This would be impossible for the visually reliant old world species. This point illustrates how the environment can impact search efficiency depending on the sensory system that's used.

Flight is also cheaper means of locomotion than running \citep{tucker1975energetic}. This advantage can be extended further in larger species by engaging in soaring instead of flapping flight, which is even cheaper still (approximately twice BMR) \citep{hedenstrom1993migration}. The advantages this confers are clear from the information we have on the enormous foraging ranges of some seabirds and accipiters \citep{bamford2007ranging}. Clearly, it would be pointless to have incredible detection abilites and not have a cost efficient movement to benefit from it. Although, as with detection ability, the environment has a role to play here. Vultures and eagles tend to soar using thermals and if these air pockets don't form, say on a cloudy day, the bird is grounded \citep{mundy1992vultures}. In many habitats (e.g. the arctic) it is simply not possible for sufficiently powerful thermals to form and as a consequence large-bodied vultures cannot exist. The upshot of this is that terrestrial carnivores like bears and wolves take more carrion \citep{devault2003scavenging}.

@@

Large body size confers substantial dominance benefits \citep{ruxton2004obligate}. Thus, we would expect scavengers to have this trait selected for even in the case of weight-constrained fliers. Cinereous Vultures (\textit{Aegypius monachus}) and condors (\textit{Vultur gryphus}, \textit{Gymnogyps californianus}) all have body masses that can exceed 10 kg and represent some of the heaviest bird species capable of flight \citep{ferguson2001raptors,donazar2002effects}. And as we have noted the Azhdarchid pterosaurs were far bigger again, with estimated body masses of around 80 kg.

@@

Similar to vultures they have well developed sensory organs, particularly in olfaction whereby they can detect a rotting carcass 2 km downwind. They have a characteristic "rocking horse gait"  which allows them to cover great distances efficiently. The bone crushing ability of hyenas reveals another useful scavenger trait. Since carrion is not dispatched directly, often the most easily accessible and choicest components of the carcass will be missing or, if present, will be fought over. Being able to extract nutrients from remnants gives the scavenger a great advantage. Osteophagy is known across a range of terrestrial carnivores. Some fat-rich mammalian bones have an energy density (6.7 kJ/g) comparable with that of muscle tissue, making skeletal remains an enticing resource \citep{brown1989study}. This ability reached its zenith among hyenas with the evolution of the 110 kg \textit{Pachycrocuta brevirostris} during the Pliocene \citep{palmqvist2011giant}. Some work on extinct sabretooths suggests they may have left a large amount of food for would-be scavengers because of their unique skull morphology. As a result the decline of Machairodontinae sabretooths has been offered as an explanation for the extinction of \textit{P brevirostris} \citep{palmqvist2011giant}. A major difficulty for terrestrial scavengers is competition with vultures. Noctural behaviour in the Hyaenidae in general has been put forth as an adaptation to reduce competition with these exclusively diurnal birds \citep{gittleman2013carnivore}.

The order Carnivora sees its origins in the Eocene where it split into the Caniforma and Feliforma. And many of the aforesaid adapations for scavenging are found in these other major terrestrial mammalian carnviores. Though the specific mix of features realised in hyenas suggest this is the model organism for terrestrial scavenging among mammals in the past. Indeed, the bone-crushing dogs that evolved during the Oligocene (subfamily Borophaginae) have been compared to hyenas in terms of their feeding ecology \citep{van2003chapter,martin2016pursuit}.

Interestingly such comparisons have given insight into the feeding ecology of early hominins who, for instance, had the ability to craft tools for breaking open bones \citep{hone2010feeding,ARCM:ARCM12084}. The question of where our ancestors placed on the hunter-scavenger axis during the Plio-Pleistocene has been a matter of debate for years. A recent study investigating potential scavenging opportunities for hominins in Kenya found that, even when discounting bone material, there is a substantial amount of scavengeable meat left on predated remains; sufficient to sustain the requirements of an adult male \textit{Homo erectus} \citep{pobiner2015new}. In some historical hominin-inhabited areas there were a greater number of felids than hyenids. This is significant because hyenas will leave far less flesh on a carcass than a felid such as a sabretooth enabling contemperaneous hominins to benefit \citep{pobiner2015new}. The intelligence, resultant tool-use and cooperative nature of hominins means they could likely adapt to take on more or less carrion depending on their environment.

By contrast to mammals, a successful reptilian scavenger requires a far different set of adapations. Modern forms are ectothermic, limiting their activity periods. This is exacerbated by the sprawling gait seen in lizards which results in Carrier's Constraint such that the animal can't move and breathe at the same time because the lateral movements impedes its lungs \citep{carrier1987evolution}. This manifests itself in aspects such as maximum sustainable speed where an equivalent mammal has a six to seven fold increase \citep{ruben1995evolution}. A lower metabolism does give reptiles an advantage however, in that over the course of a year their food requirements can be 30 times smaller than an endotherm of equal size \citep{Nagy1621}. Any adaptations that reduce energetic costs are likely to be selected in scavengers. \cite{devault2002scavenging} suggest this is an avenue for scavenging in snakes because they "exhibit  exceedingly  low  maintenance  metabolisms,  and most  can  survive  on  a  few  scant  feedings per year. It  is, therefore, possible for snakes to rely largely  on  infrequent,  less  energy-rich  meals." In the same review the authors found occurrences of scavenging spread across five families of snakes and stated that this behaviour is "far more common than currently acknowledged."\citep{devault2002scavenging}.

Unsurprisingly, given their enduring appeal, the prevalence of scavenging has been explored in the carnivorous, theropod dinosaurs. These animals ranged from the chicken-sized to the whale-sized all of which were bipedal. They are quite alien to anything we know today which restricts our ability to understand their ecology far more so than extinct mammals \citep{weishampel2004dinosauria}. Of relevance, are the questions that still persist about their metabolism \citep{grady2014evidence} and sensory perception. We do know that they walked with the erect gait of mammals or birds rather than the sprawling gait of lizards and that they were most likely facultative scavengers \citep{depalma2013physical}. Much work has focused on the existence of the behaviour in \textit{Tyrannosaurus rex} \citep{ruxton2003could,carbone2011intra} but a recent energetics study investigated the likely prevelance of scavenging across a range of body sizes. In it the authors demonstrated that species of intermediate body masses (approx. 500 kg) would have gained the most benefit from scavenging. This was the result of gut capacity limitations and the effects of competition at the carcass. At the larger extreme this owes to the fact that gut capacity doesn't scale isometrically with body mass so the benefits of greater mass level off; there's only so much food an individual can consume at a single sitting. For the smaller species, larger competitors would have prevented their access to carrion.

@@

Of course, tetrapod terrestrial dominance predates the evolution of the dinosaurs. It is in the Permian, almost 300 millions years ago, that we have the earliest evidence of vertebrate scavenging where a temnospondyl amphibian fed on the carcass of \textit{Varanops}, a predatory synapsid of the time \citep{reisz2006articulated}. The absence of flying vertebrates in the Palaeozoic may have permitted terrestrial forms to take in a higher proportion of carrion in their diet. As noted earlier, a system similar to the Serengeti could support a terrestrial scavenger \citep{ruxton2004obligate}

\section*{Aquatic Scavengers} 
An aquatic environment presents challenges for direct observational studies and so, similar to the approaches involving extinct species, much work has approached the question of scavenging propensity from an energetics perspective. Although as with the aerial and terrestrial enviornments we have evidence of facultative scavenging among aquatic species. For example, the remains of a mosasaur and a terrestrial hadrosaur were discovered with embedded teeth from a Cretaceous shark \textit{Squalicorax} \citep{schwimmer1997scavenging}. As well as a likely instance of scavenging between a 4-million-year-old white shark (\textit{Carcharodon}) and mysticete whale from Peru \citep{ehret2009caught}.
The existence of an obligate scavenger in a marine setting is uncertain \citep{britton1994marine,smith2003ecology,ruxton2004energetic,ruxton2005searching}. Depending on the species a carcass in this environment either floats or descends to the sea floor. In this latter low-light environment detection distances are far lower (< 100 m) than they would be in the air. As such, animals detect resources through chemo- and mechanoreception more so than through vision \citep{ruxton2004energetic}. However, water is a medium that is conducive to low-cost movement \citep{tucker1975energetic} and so may be able to support an obligate scavenging fish \citep{ruxton2004energetic,ruxton2005searching}. \cite{benbow2015introduction} do note that "some benthic scavengers (e.g., hagfish: family Myxinidae) rely on necrophagy for a large portion of their diet and may indeed be obligate scavengers".

@@

 %\citep{sazima1990necrofagia} predicted that aquatic or semi-aquatic piscivorous snakes  would  scavenge  more  frequently than other species. Moreover, \cite{sazima1990necrofagia}  suggested that water currents that induce  aggregations of carrion heighten the probability of carrion detection by these snakes (see also \cite{savitzky1992laboratory}).  Also,  chemical  gradients may  be  more  uniform  and  give  more  de-pendable  directional information in water \citep{sazima1990necrofagia}." \citep{devault2002scavenging}

>>>>

\section*{Introduction}
%TG: it might be better to reorganise the intro as:
%§1 - scavenging is understudied because it's hard to say who's a scavenger
%§2 - it's even more problematic in the fossil record but people have tried some indirect approaches
%§3 - good thing is that we can apply these indirect approaches to extant systems
%§4 - in this review we discuss some stuff.
%TG: or something along these lines, obviously flexible!
Historically, scavengers have not been viewed as the most charismatic of animals. This may go some way to explaining the gap in our knowledge of the prevalence of this behaviour. Consider Professor Sanborn Tenney writing in 1877 for The American Naturalist who had this to say about one well known group, “Prominent among the mammalian scavengers are the hyenas, the ugliest in their general appearance of all the flesh eaters.” He contrasts these with “nobler kinds” of carnivores such as lions and tigers \citep{tenney1877few}. Even aside from our own subjective biases, scavenging is a difficult behaviour to detect after the fact. Without catching a carnivore in the act of killing we are left to infer how the prey was killed. Some simple heuristics can inform us, for instance, in cases where the prey item was simply too large to have been killed by the ostensible predator \citep{pobiner2008paleoecological}. But clearly, a scavenger doesn’t only feed on animals too big for it to have hunted. The obvious lack of direct behavioural data compounds the difficulty of discerning scavenging among extinct forms. Indeed, a single species of dinosaur notwithstanding, a synthesis describing the natural history of scavengers is absent from the literature. Fortunately, research on scavenging is on the rise \citep{manga2006vulture}. As a result, we are now beginning to realise the extent of this behaviour such that, “in some ecosystems, vertebrates have been documented to assimilate as much as 90\% of the available carrion" \citep{benbow2015introduction}. Even Tenney’s noble big cats are now known to take in a significant portion of carrion in their diet where some lion populations get over 50\% of their meat from carcasses. By recognising the difficulty in directly observing scavenging, a suite of methods have been used to discern the most suitable morphologies, physiologies and environments for a scavenging lifestyle to prosper. Here we chart the natural history of vertebrate scavenging on air, land and in water by focussing on the dominant vertebrate groups. %It is our aim in this review to employ these methods to gain an understanding of scavengers past and present. 
\subsection*{The Difficulty of Scavenging}
The chief hurdle to scavenging is finding a sufficient quantity of food, the occurence of which is difficult to predict in space and time. Once found the scavenger has to then overcome the agents of decay produced by the action of microorganisms on the carcass \citep{ruxton2014fruit}. The idea of scrounging from predator kills is undermined by studies showing that in the  majority of ecosystems more animals die from disease and starvation than predation \citep{benbow2015introduction}. Thus, any animal existing as a scavenger must maximise its detection capabilities and minimise its locomotory costs \citep{ruxton2004obligate}. The habitat must also be productive enough to sustain an animal biomass that will eventually produce carcasses.

@@

Vultures represent the best known scavengers on Earth. These birds consist of two convergent groups, from the old and the new world and represent the only example of obligate vertebrate scavengers today. Given their unique position, they have been extensively studied to determine what adaptations they possess that allows them to so flourish in this niche. Underscoring their prowess are studies such as \cite{ogada2012effects} who note a three fold increase in carcass persistence times in the absence of vultures. As such, we can begin by exploring the adapdations and the environments of vultures to draw comparisons with other scavenging species and \textit{their} environments.

@@

Aside from sight, three species within the new world family Cathartidae, (genus \textit{Cathartes}), have well developed olfactory systems \citep{AR:AR22815}. Among them are the Turkey Vultures (\textit{Cathartes aura}) which were able to locate 90\% of baits set out in a tropical forest \citep{houston1986olfaction}. This would be impossible for the visually reliant old world species. This point illustrates how the environment can impact search efficiency depending on the sensory system that's used.

Flight is also a cheaper means of locomotion than running \citep{tucker1975energetic}. This advantage can be extended further in larger species by engaging in soaring instead of flapping flight, which is even cheaper still (approximately twice BMR) \citep{hedenstrom1993migration}. The benefits this confers are clear from the information we have on the enormous foraging ranges of some seabirds and accipiters \citep{bamford2007ranging}. Clearly, it would be pointless to have incredible detection abilites and not have a cost efficient movement to benefit from it. Although, as with detection ability, the environment has a role to play here. Vultures and eagles tend to soar using thermals and if these air pockets don't form, say on a cloudy day, the bird is grounded \citep{mundy1992vultures}. In many habitats (e.g. the arctic) it is simply not possible for sufficiently powerful thermals to form and as a consequence large-bodied vultures cannot exist. The upshot of this is that terrestrial carnivores like bears and wolves take more carrion \citep{devault2003scavenging}.

@@

Large body size confers substantial dominance and starvation resistance benefits \citep{ruxton2004obligate}. Thus, we would expect scavengers to have this trait selected for even in the case of weight-constrained fliers. Cinereous Vultures (\textit{Aegypius monachus}) and condors (\textit{Vultur gryphus}, \textit{Gymnogyps californianus}) all have body masses that can exceed 10 kg and represent some of the heaviest bird species capable of flight \citep{ferguson2001raptors,donazar2002effects}. And as we have noted the Azhdarchid pterosaurs were far bigger again, with estimated body masses of around 80 kg.

@@

Similar to vultures they have well developed sensory organs, particularly in olfaction whereby they can detect a rotting carcass 2 km downwind. They have a characteristic "rocking horse gait"  which allows them to cover great distances efficiently. A major difficulty for terrestrial scavengers is competition with vultures. Noctural behaviour in the Hyaenidae in general has been put forth as an adaptation to reduce competition with these exclusively diurnal birds \citep{gittleman2013carnivore}. The bone crushing ability of hyenas reveals another useful scavenger trait. Since carrion is not dispatched directly, often the most easily accessible and choicest components of the carcass will be missing or, if present, will be fought over. Being able to extract nutrients from remnants gives the scavenger a great advantage. Osteophagy is known across a range of terrestrial carnivores. Some fat-rich mammalian bones have an energy density (6.7 kJ/g) comparable with that of muscle tissue, making skeletal remains an enticing resource \citep{brown1989study}. This ability reached its zenith among hyenas with the evolution of the 110 kg \textit{Pachycrocuta brevirostris} during the Pliocene \citep{palmqvist2011giant}. Some work on extinct sabretooths suggests they may have left a large amount of food for would-be scavengers because of their unique skull morphology. As a result the decline of Machairodontinae sabretooths has been offered as an explanation for the extinction of \textit{P brevirostris} \citep{palmqvist2011giant}.

The order Carnivora sees its origins in the Eocene where it split into the Caniforma and Feliforma. And many of the aforesaid adapations for scavenging are found in these other major terrestrial mammalian carnviores. Though the specific mix of features realised in hyenas suggest this is the model organism for terrestrial scavenging among mammals in the past. For example, the bone-crushing dogs that evolved during the Oligocene (subfamily Borophaginae) have been compared to hyenas in terms of their feeding ecology \citep{van2003chapter,martin2016pursuit}. Certainly, it is profitable to compare extant terrestrial species to their prehistoric forebears given the dominance of mammalian carnviores since the Eocene.

Interestingly such comparisons have given insight into the feeding ecology of early hominins who, for instance, had the ability to craft tools for breaking open bones \citep{hone2010feeding,ARCM:ARCM12084}. The question of where our ancestors placed on the hunter-scavenger axis during the Plio-Pleistocene has been a matter of debate for years. A recent study investigating potential scavenging opportunities for hominins in Kenya found that, even when discounting bone material, there is a substantial amount of scavengeable meat left on predated remains; sufficient to sustain the requirements of an adult male \textit{Homo erectus} \citep{pobiner2015new}. In some historical hominin-inhabited areas there were a greater number of felids than hyenids. This is significant because hyenas will leave far less flesh on a carcass than a felid such as a sabretooth enabling contemperaneous hominins to benefit \citep{pobiner2015new}. The intelligence, resultant tool-use and cooperative nature of hominins meant they could likely adapt to take on more or less carrion depending on their environment.

By contrast to mammals, a successful reptilian scavenger requires a far different set of adapations. Modern forms are ectothermic, limiting their activity periods. This is exacerbated by the sprawling gait seen in many lizards which results in Carrier's Constraint such that the animal can't move and breathe at the same time because the lateral movements impedes its lungs \citep{carrier1987evolution}. This manifests itself in aspects such as maximum sustainable speed where an equivalent mammal has a six to seven fold increase \citep{ruben1995evolution}. A lower metabolism does give reptiles an advantage however, in that over the course of a year their food requirements can be 30 times smaller than an endotherm of equal size \citep{Nagy1621}. Any adaptations that reduce energetic costs are likely to be selected in scavengers. \cite{devault2002scavenging} suggest this is an avenue for scavenging in snakes because they "exhibit  exceedingly  low  maintenance  metabolisms,  and most  can  survive  on  a  few  scant  feedings per year. It  is, therefore, possible for snakes to rely largely  on  infrequent,  less  energy-rich  meals." In the same review the authors found occurrences of scavenging spread across five families of snakes and stated that this behaviour is "far more common than currently acknowledged."\citep{devault2002scavenging}.

Unsurprisingly, given their enduring appeal, the prevalence of scavenging has been explored in the carnivorous, theropod dinosaurs. These animals ranged from the chicken-sized to the whale-sized all of which were bipedal. They are quite alien to anything we know today which restricts our ability to understand their ecology far more so than extinct mammals \citep{weishampel2004dinosauria}. Of relevance, are the questions that still persist about their metabolism \citep{grady2014evidence} and sensory perception. We do know that they walked with the erect gait of mammals or birds rather than the sprawling gait of lizards and that they were most likely facultative scavengers \citep{depalma2013physical}. Much work has focused on the existence of the behaviour in \textit{Tyrannosaurus rex} \citep{ruxton2003could,carbone2011intra} but a recent energetics study investigated the likely prevelance of scavenging across a range of body sizes. In it the authors demonstrated that species of intermediate body masses (approx. 500 kg) would have gained the most benefit from scavenging. This was the result of gut capacity limitations and the effects of competition at the carcass. At the larger extreme this owes to the fact that gut capacity doesn't scale isometrically with body mass so the benefits of greater mass level off; there's only so much food an individual can consume at a single sitting \citep{calder1996size}. For the smaller species, larger competitors would have prevented their access to carrion.

@@

Of course, tetrapod terrestrial dominance predates the evolution of the dinosaurs. It is in the Permian, almost 300 millions years ago, that we have the earliest evidence of vertebrate scavenging where a temnospondyl amphibian fed on the carcass of \textit{Varanops}, a predatory synapsid of the time \citep{reisz2006articulated}. The absence of flying vertebrates in the Palaeozoic may have permitted terrestrial forms to take in a higher proportion of carrion in their diet. As noted earlier, a system similar to the Serengeti in productivity could have supported a terrestrial scavenger \citep{ruxton2004obligate}.

\section*{Aquatic Scavengers} 
An aquatic environment presents challenges for direct observational studies and so, similar to the approaches involving extinct species, much work has approached the question of scavenging propensity from an energetics perspective. Although as with the aerial and terrestrial enviornments we have evidence of facultative scavenging among aquatic species. For example, the remains of a mosasaur and a terrestrial hadrosaur were discovered with embedded teeth from a Cretaceous shark \textit{Squalicorax} \citep{schwimmer1997scavenging}. As well as a likely instance of scavenging between a 4-million-year-old white shark (\textit{Carcharodon}) and mysticete whale from Peru \citep{ehret2009caught}.
The existence of an obligate scavenger in a marine setting is uncertain \citep{britton1994marine,smith2003ecology,ruxton2004energetic,ruxton2005searching}. Depending on the species a carcass in this environment either floats or descends to the sea floor. In the latter low-light environment, detection distances are far lower (< 100 m) than they would be in the air. As such, animals detect resources through chemo- and mechanoreception more so than through vision \citep{ruxton2004energetic}. However, water is a medium that is conducive to low-cost movement \citep{tucker1975energetic} and so may be able to support an obligate scavenging fish \citep{ruxton2004energetic,ruxton2005searching}. \cite{benbow2015introduction} do note that "some benthic scavengers (e.g., hagfish: family Myxinidae) rely on necrophagy for a large portion of their diet and may indeed be obligate scavengers".

@@

\section*{The Role of Climate}
The geological record shows the Earth has undergone radical fluctuations in temperature. This will have had a significant bearing on the availability of carrion. A 10$^{\circ}$C increase in ambient temperature can double carcass decomposition rates \citep{benbow2015introduction}. Oceanic productivity, decomposition rates and habitat structure are all impacted by climactic conditions. \cite{benbow2015introduction} points to a series of studies showing how microbes and invertebrates benefit at higher temperatures to the detriment of vertebrate scavengers "above 20$^{\circ}$C vertebrates were able to detect and consume only 19\% of small-mammal carcasses, whereas at temperatures below 18$^{\circ}$C, vertebrates consumed 49\% of carcasses". Furthermore, "carcass decomposition occurs more rapidly in warm, moist habitats when compared with more xeric habitats."

% Slight changes in temperature could have profound impacts on energy flow within ecosystems as carrion decomposition
% rates roughly double for every 10°C increase in ambient temperature (Vass et al. 1992; Parmenter and
% MacMahon 2009). Consequently, the availability of carrion to vertebrate scavengers could be reduced by
%as much as 20%–40% over the next century based on projections from current climate change models
% (Beasley et al. 2012b).

%"Changes in global climate are also expected to substantially alter the availability of carrion resources in marine ecosystems as surface production of organic material could decline by 50\% or more (Smith et al. 2008). Such a drastic reduction in productivity undoubtedly will impact a multitude of ecosystem processes, including the availability and distribution of carrion within marine ecosystems." \citep{benbow2015introduction}

%"For example, vertebrate scavengers appear to be disadvantaged when humidity and temperature favor microbial and invertebrate reproduction (between 21C and 38C)" \citep{benbow2015introduction}

%"In a study by DeVault et al. (2004b), it was found that at temperatures above 20C vertebrates were able to detect and consume only 19\% of small-mammal carcasses, whereas at temperatures below 18C, vertebrates consumed 49\% of carcasses" \citep{benbow2015introduction}

%"For example, the available literature suggests that carcass decomposition occurs more rapidly in warm, moist habitats when compared with more xeric habitats" \citep{benbow2015introduction}

%Of particular note is the increase in decomposition rate of carrion with temperature which could result in a seasonal reduction of carrion availability given predicted warming trends \citep{moleon2014inter}.
